---
title: Assignment 5: Evaluating the impact of PCVs on pneumonia hospitalization in the US
output:
  rmdformats::readthedown:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: false
    css: ./Data/stylesreadthedown.css
    toc_depth: '3'
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to the Assignment

You have been asked to evaluate the impact of the introduction of pneumococcal conjugate vaccine (PCV7) in the US. PCV7 was introduced for use in infants in the US in March 2000, with a 4 dose schedule (doses at 2,4,6,12 months of age). These data come from hospitalizations from 8 large states (combined together). 

Your task is to write a 1 page 'executive summary' of the impact of PCVs on hospitalization for pneumonia in infants (2-11 month olds), as well as possible indirect effect of the vaccine on older adults (age 80+ years). 
- *The target audience for this is the ACIP Working group on Pneumococcal Vaccination*. This is a group of subject matter experts including individuals from the CDC, academia, and medicine who evaluate evidence and provide guidance on vaccine policy to the broader ACIP. You can include plots as needed, but be strategic.

- For both age groups, you should evaluate the output from 2 types of analyses: an interrupted time series analysis and a synthetic control-type analysis (fit to pre-vaccine data and include control time series)  You should mention the agreement/disagreement between methods in your summary and which specific variables/parameters you used.

- You should also discuss the strength of the evidence. Your judgement on this can be influenced by the uncertainty intervals of the estimates, by concordance/discordance between methods, by univariate analyses with different controls, and by prior expectation of what a reasonable range of results would be

-As a denominator, you could consider select 'all_cause_no_resp' (all cause hospitalizations, excluding those with a respiratory cause)

*You can work in groups of 1-3 people for this assignment (you only need to hand in 1 assignment per group on canvas; please write all the names on the assignment)*

The code below is from the in-person exercise and needs to be modified to find an optimal analysis.


### Load the libraries we need 

```{r}
library(dplyr)
library(ggplot2)
library(MASS)
library(lubridate)
library(plotly)
source('./R/its_outputs.R')
source('./R/control_outputs.R')

source('./R/scale.fun.R')
source('./R/run_its.R')
source('./R/run_control.R')

```

## Read in the data
```{r}
a2 <- readRDS('./Data/US.rds') %>%
  select(Pneum, date, age,all_cause_no_resp, starts_with('X_') ) %>%
  filter(date <= '2005-12-01') %>%
  filter(age=='2-11m') #select the age group here
```

Set parameters here for the model
```{r}
vax.intro.date= as.Date('2000-01-01')

months_until_second_knot = 12 #how many months between vaccine intro and start of evaluation period

months_until_third_knot = 48  #if there is a 2nd change in slope, how many months after vaccine intro does it occur? if no 2nd change, set as 999

```




```{r}
form_its1 <- as.formula ("Pneum ~ month+ index + spl1 + spl2   +spl3 + offset(all_cause_no_resp)")


mod1 <- run_its(set.vax.intro.date= vax.intro.date, set.form_its1=form_its1, months_until_third_knot=months_until_third_knot, months_until_second_knot=months_until_second_knot )

summary(mod1$mod1)

mod1$mod1$aic #AIC score

```

Run this function to generate uncertainty intervals and summary stats for the ITS model

```{r}
mod1.outputs <- its_outputs(mod1)

```

Ratio for entire evaluation period

```{r}
mod1.outputs$rr.q.post
```

Rate ratio over time from ITS
```{r}
mod1.outputs$p.rr.trend +ggtitle('Rate Ratio over time following vaccine introduction')
```

Observed vs expected by month. Observed=red; black=fitted; gray shaded area=counterfactual
```{r}
mod1.outputs$p.preds
```

Observed vs expected by year Observed=red; black=fitted; gray shaded area=counterfactual
```{r}
mod1.outputs$p.preds.agg
```

Cumulative cases prevented
```{r}
ggplotly(mod1.outputs$p.cum_prevented)
```


## Model 2: Fit to pre-period, and project forward (similar to synthetic controls)

Note you could adjust for trend (with index) or not here; it is debatable what the best strategy is. Just be consistent and justify your approach, and specify it a priori.

```{r}

form_control <- as.formula ("Pneum_pre ~ month+ X_001_139_SY + X_240_279+ X_280_289 + X_320_359_SY + X_390_459 + X_520_579 + all_cause_no_resp+ X_580_629 + X_680_709 + X_780_799 + X_800_999 + X_V00_V91")

mod2 <- run_control(set.vax.intro.date= vax.intro.date, set.form_cont1=form_control)

summary(mod2$mod1)

mod2$mod1$aic #AIC score (note cannot be compared with ITS model because they are fit to different sets of Y variables)

```
format the outputs, generate plots

```{r}
mod2.outputs <- control_outputs(fitted.mod=mod2,months.start=months_until_second_knot,
                                set.vax.intro.date=vax.intro.date,set.months_until_second_knot=months_until_second_knot)
```

Rate ratio for eval period

```{r}
mod2.outputs$rr.q.post

```

Rate ratio at each time points

```{r}
mod2.outputs$p.rr.trend
```

Observed vs expected by month
```{r}
mod2.outputs$p.preds
```

Observed vs expected by year
```{r}
mod2.outputs$p.preds.agg
```

```{r}
ggplotly(mod2.outputs$p.cum_prevented)
```
